\section{Small Language Models in Legal Domain}

\subsection{Opportunities and Constraints}

The deployment of Large Language Models (LLMs) in practical legal workflows presents severe operational challenges. As discussed in Chapter 1, proprietary API-dependent models pose risks to data confidentiality under national guidelines, and incur prohibitive operating costs when processing long statutory texts. Consequently, Small Language Models (SLMs) --- generally defined as architectures with fewer than 3 billion parameters --- have emerged as a viable alternative. Local on-premise deployment of SLMs guarantees absolute data privacy and reduces computational latency. However, the primary constraint of SLMs is their restricted parameter capacity, which severely limits their multi-step reasoning depth and increases their susceptibility to hallucinations.

\subsection{Survey of Legal Adaptation Frameworks}

To bridge the performance gap between lightweight models and frontier LLMs, several specialized legal adaptation and knowledge distillation frameworks have been proposed:

First, LegalDrill \parencite{li2026legaldrilldiagnosisdrivensynthesislegal} introduced a diagnosis-driven synthesis framework for legal reasoning in Small Language Models. LegalDrill evaluates the student model's output using an internal logits-based difficulty score, identifying high-uncertainty prompts, and then uses a teacher model to synthesize targeted training data to correct specific diagnostic errors. Second, Italiani et al. \parencite{italiani2025enhancing} proposed a legal question answering framework that utilizes large language models for dataset generation and subsequent token-level knowledge distillation to lightweight student models. Their work demonstrated that soft probability distributions (logits) transfer implicit domain knowledge more effectively than hard-label fine-tuning alone. 

Third, in the context of the Vietnamese legal domain, benchmarks such as the LegalSLM Shared Task \parencite{le-etal-2025-overview} and VLegal-Bench \parencite{dong2026vlegalbenchcognitivelygroundedbenchmark} have been established to systematically evaluate the reasoning capabilities of small models. Evaluations on these benchmarks indicate that general-domain SLMs (such as the base Qwen and Llama architectures under 3B parameters) exhibit high error rates when parsing statutory structures and generating step-by-step explanations, highlighting the need for domain-specific distillation pipelines.

\subsection{Research Gaps in Legal Distillation}

While existing frameworks have improved legal task performance, several critical research gaps remain unaddressed:
\begin{enumerate}
    \item \textbf{Exposure Bias in Sequential Reasoning:} Standard distillation methods (such as those in \parencite{italiani2025enhancing} and standard SFT) rely on off-policy teacher-forcing. They do not address the exposure bias problem during multi-step logical deduction (such as syllogistic explanation generation), where early logical mistakes compound quadratically along the student's own generative trajectory.
    \item \textbf{Miscalibration and Difficulty Estimation:} Selection mechanisms that rely on internal logits-based difficulty scores (such as in LegalDrill) are highly unreliable when applied to extremely small models (under 3 billion parameters). Due to restricted capacity, these SLMs suffer from severe calibration bias, often assigning high confidence to incorrect reasoning chains and hallucinations, thereby bypassing logits-based filtering.
    \item \textbf{Jurisdictional and Structural Specificity:} Existing distillation methods are primarily designed for English or Chinese legal contexts. They lack optimization for the highly nested, article-by-article logical structures characteristic of the Vietnamese Civil Law system, which requires strict alignment across major premises, minor premises, and conclusions.
\end{enumerate}
This study addresses these gaps by proposing a three-phase distillation and preference alignment framework specifically optimized for Vietnamese legal reasoning.
