\section{Reinforcement Learning for Legal Reasoning}

\subsection{Group Relative Policy Optimization}

Reinforcement Learning from Human Feedback (RLHF) has become the standard paradigm for aligning Large Language Models (LLMs) with human preferences and enhancing their complex reasoning capabilities. Traditionally, algorithms such as Proximal Policy Optimization (PPO) require the deployment of both an actor policy and a critic model, where the critic model is trained to estimate state-value functions as baselines. However, maintaining a separate critic model significantly increases GPU memory footprint and computational latency, particularly as the base model scales.

To mitigate these resource constraints, Group Relative Policy Optimization (GRPO) has been proposed as a critic-free reinforcement learning algorithm. Instead of using a critic model to estimate absolute values, GRPO evaluates the quality of generated sequences by comparing them relatively within a sampled group. Specifically, for each input prompt $q$, the active policy $\pi_\theta$ samples a group of $G$ independent rollouts $\{o_1, o_2, \dots, o_G\}$. A reward model (or a set of verifiable rule-based checkers) computes a reward $r_i$ for each rollout $o_i$. The relative advantage $A_i$ of rollout $o_i$ is then computed using the group mean and standard deviation:
\begin{equation}
    A_i = \frac{r_i - \frac{1}{G}\sum_{j=1}^G r_j}{\text{std}(r_1, \dots, r_G)}.
\end{equation}
The policy objective of GRPO is to maximize:
\begin{align}
    \mathcal{L}_{\text{GRPO}}(\theta) = \frac{1}{G} \sum_{i=1}^G \Bigg[ &\min\left(
    \frac{\pi_\theta(o_i \mid q)}{\pi_{\theta_{\text{old}}}(o_i \mid q)} A_i,
    \text{clip}\left(\frac{\pi_\theta(o_i \mid q)}{\pi_{\theta_{\text{old}}}(o_i \mid q)}, 1-\epsilon, 1+\epsilon\right) A_i
    \right) \notag\\
    &{} - \beta D_{KL}(\pi_\theta \parallel \pi_{\text{ref}}) \Bigg],
\end{align}
where $\pi_{\text{ref}}$ is a frozen reference policy, $\epsilon$ is the clipping threshold, and $\beta$ is the Kullback-Leibler (KL) divergence penalty coefficient that restricts the active policy from drifting too far from the reference policy. 

In legal reasoning tasks (such as syllogistic deduction), where verification rules are strict, GRPO enables the policy to learn optimal structural and logical reasoning paths. By optimizing the model on verifiable rewards --- such as matching specific tags like \texttt{<think>}, \texttt{<major\_premise>}, \texttt{<minor\_premise>}, and \texttt{<conclusion>} --- the teacher model acquires strong reasoning behaviors without the need for resource-intensive value network training.

\subsection{Privileged Information in Distillation}

The paradigm of Learning Under Privileged Information (LUPI) describes a training scenario where a teacher model has access to additional features or context, denoted as $x^*$ (privileged information), that are hidden from the student model during training and are entirely unavailable during inference. 

In on-policy distillation frameworks, the teacher model acts as an evaluator of the student's self-generated trajectories. To ensure that the teacher's feedback is highly accurate and authoritative, the teacher model is operated in an "Oracle" mode. In this mode, the teacher model is provided with the ground-truth legal answer and context as privileged information $x^*$. When the student model generates a reasoning rollout $o$ for a prompt $q$, the teacher model computes its target probability distribution by conditioning its forward pass on both the prompt and the privileged gold standard:
\begin{equation}
    P_T(\cdot \mid q, o, x^*).
\end{equation}
Conditioning on the gold answer allows the teacher model to evaluate the correctness, logical validity, and formatting of the student's rollout with near-absolute confidence. The resulting soft targets generated by the teacher contain high-quality, targeted supervision signals. During distillation, the student model, which is parameterized to predict the next token given only the input prompt $q$, is trained to minimize the divergence between its output distribution $P_S(\cdot \mid q, o)$ and the teacher's privileged distribution:
\begin{equation}
    \mathcal{L}_{\text{oracle}} = D_{KL}\left( P_T(\cdot \mid q, o, x^*) \parallel P_S(\cdot \mid q, o) \right).
\end{equation}
By distilling this privileged distribution, the student model inherits the reasoning logic of a highly confident oracle, allowing the student to learn correct logical paths while operating under standard inference conditions where the gold answers are absent \parencite{li2026legaldrilldiagnosisdrivensynthesislegal, zhao2026trainingharnessesonpolicyharness}.
