\section{Limitations}

Despite the promising results reported above, several limitations qualify the scope of our conclusions and motivate the directions outlined in the next section.

\begin{enumerate}
    \item \textbf{Capacity-induced instability at the 0.6B scale.} The on-policy and preference-optimization phases were counterproductive for the smallest student. The 0.6B model's Overall score fell to 0.3104 after On-Policy Knowledge Distillation and collapsed further to 0.1861 after DPO, indicating that the probability-displacement penalty and the noise of self-generated rollouts exceed the absorptive capacity of a sub-billion-parameter network.
    \item \textbf{Over-specialization and degraded out-of-distribution generalization.} On the external VLSP2025 Public-Test, the 1.7B SFT baseline achieved the highest Overall score (0.6395), surpassing all distilled and aligned variants. This reveals that knowledge distillation, while effective in-distribution, over-specializes the student toward the training distribution at the expense of generalization. The effect is most acute on the MCQ task, where every distilled model scored below 0.56 against the SFT baseline's 0.74.
    \item \textbf{Proxy evaluation metric for syllogistic reasoning.} For the Syllogism task we reported ROUGE-L as an internal proxy, whereas the official VLSP2025 protocol employs an LLM-as-a-Judge. The reported syllogism figures therefore reflect lexical overlap rather than the official competition scoring and may not fully capture deductive correctness.
    \item \textbf{Narrow experimental scope.} The study relied on a single training corpus of 2,603 samples within one legal domain (Vietnamese statutory law) and a single teacher model (\texttt{Qwen3-4B-legal-GRPO}). The student's attainable performance is consequently bounded by the quality and coverage of this individual teacher.
    \item \textbf{Dependence on an external LLM judge.} The diagnosis-driven preference data was generated using a proprietary external model (\texttt{GPT-4o-mini}), which introduces monetary cost and limits the full reproducibility and auditability of the alignment data pipeline.
    \item \textbf{Hardware and efficiency constraints.} Owing to limited compute, we did not evaluate larger students (4B and beyond), and the on-policy phase remained computationally expensive, running roughly eight times slower than offline distillation.
\end{enumerate}
