\section{Overview of the Proposed Pipeline}

This chapter details the proposed methodology for distilling advanced legal reasoning capabilities from a large, reinforcement-learning-aligned Teacher Large Language Model (LLM) into a lightweight Student Small Language Model (SLM). The primary goal is to produce a small model that exhibits robust, logically sound, and structurally consistent reasoning paths without incurring high computational overhead or violating data privacy regulations.

\subsection{Problem Formulation for Legal Reasoning}

We formulate the task of legal reasoning as a conditional sequence generation problem. Let $x = (c, q) \in \mathcal{X}$ denote the input instance, where $c \in \mathcal{C}$ represents the legal context (such as statutory texts, constitutional provisions, or relevant case files) and $q \in \mathcal{Q}$ represents the query (such as the factual description of a case or a specific legal question). 

The target output is a structured sequence $y \in \mathcal{Y}$. In generative reasoning models, the output sequence $y$ is defined as a tuple $y = (s, v)$, where $s = (s_1, s_2, \dots, s_{|s|})$ is the intermediate reasoning chain (the "thinking process" or chain-of-thought) and $v = (v_1, v_2, \dots, v_{|v|})$ represents the final verdict or answer. The reasoning chain $s$ details the logical deduction steps, such as the extraction of the major premise, the application of the minor premise, and the syllogistic reasoning. The verdict $v$ corresponds to the final concrete decision (e.g., a classification label such as \texttt{relevant}/\texttt{not\_relevant}/\texttt{unknown} or a specific article violation).

The objective is to model the conditional probability distribution $P(y \mid x)$ over a token vocabulary $\mathcal{V}$. Auto-regressively, this is defined as:
\begin{equation}
    P(y \mid x) = \prod_{t=1}^{|y|} P(y_t \mid y_{<t}, x),
\end{equation}
where $y_t \in \mathcal{V}$ is the token generated at step $t$, and $y_{<t} = (y_1, y_2, \dots, y_{t-1})$ represents the prefix history of generated tokens. 

Our goal is to optimize the parameters $\theta$ of a Student model $\pi_\theta(y \mid x)$ to approximate the target distribution defined by a Teacher model $\pi_T(y \mid x)$, while ensuring that $\pi_\theta$ remains computationally lightweight and highly accurate in its logical outputs.

\subsection{Overall Architecture of the Framework}

To transfer structured legal reasoning capabilities effectively while addressing constraints like exposure bias and logical hallucinations, we propose a three-phase optimization framework organized into two independent distillation tracks that share only the supervised fine-tuned (SFT) base. The \emph{offline track} applies Phase 1 (offline logit distillation) directly to the SFT model, and then applies Phase 3 (diagnostic preference alignment) on top of the resulting Phase 1 checkpoint---which also serves as the reference policy for the preference optimization. The \emph{on-policy track} applies Phase 2 (on-policy correction) directly to the same SFT base, independently of Phase 1. Rather than forming a single linear chain, the two tracks branch in parallel from the SFT initialization and are evaluated as alternative refinement strategies. Figure~\ref{fig:overall_pipeline} illustrates the overall architecture of the framework, which is structured as follows:

\begin{enumerate}
    \item \textbf{Phase 1: Offline Logit Distillation (Knowledge Acquisition):}
    The framework initializes the student model using a supervised anchor. The student policy $\pi_\theta$ is trained on a static training corpus using a combined loss function that optimizes both hard-label cross-entropy and soft-label Kullback-Leibler (KL) divergence against pre-computed teacher logits. This phase builds the foundational vocabulary and domain-specific linguistic knowledge of the student model.
    
    \item \textbf{Phase 2: On-Policy Knowledge Distillation (Exposure Bias Mitigation):}
    To bridge the distribution mismatch between training and inference (exposure bias), this phase is applied directly to the SFT base---independently of Phase 1---in an on-policy setting. The student model generates its own reasoning trajectories $\hat{y} \sim \pi_\theta(\cdot \mid x)$. The teacher model evaluates these student-generated rollouts. To maximize teacher confidence, the teacher operates in an Oracle mode with privileged access to the ground-truth answers. The student's parameters are updated using a token-level clipped KL divergence loss, which prevents structural style tokens from dominating the gradient updates.
    
    \item \textbf{Phase 3: Diagnosis-Driven Preference Optimization (Logical Alignment):}
    Initialized from the Phase 1 (offline) checkpoint---which also serves as the DPO reference policy---this phase continues the offline track and targets specific reasoning errors (such as logical leaps or statutory misinterpretations). An external auditor agent evaluates on-policy rollouts generated by the student, classifying errors to construct a database of failures. These failures are used to synthesize preference pairs (Chosen vs. Rejected responses). We then apply Direct Preference Optimization (DPO) to align the student model, systematically suppressing the probability of generating flawed reasoning chains while reinforcing logically consistent ones.
\end{enumerate}

\input{figs/tikz/fig_3_1_overall_pipeline}

This multi-phase pipeline systematically transforms the student model from a passive imitator into a robust, self-correcting reasoner tailored for the Vietnamese legal domain.
