\section{Phase 1: Offline Logit Distillation}

The first phase of our proposed pipeline focuses on baseline knowledge acquisition. In this phase, the student model is trained offline to emulate the token-level output distribution (logits) of the teacher model on a static training corpus. By doing so, the student model absorbs the basic linguistic patterns, domain-specific terminology, and general formatting requirements of the legal reasoning task.

\subsection{Supervised Training with Anchor}

A major challenge in knowledge distillation for generative language models is ensuring that the student model remains anchored to the ground-truth answers (hard labels) while learning from the teacher's soft probability distributions. If the student model is trained solely on soft labels, it may optimize for secondary or noisy predictions from the teacher, potentially leading to a degradation in its ability to output the exact correct verdict.

To address this challenge, we incorporate a hard-label supervised fine-tuning (SFT) anchor. The training objective is formulated as a multi-task loss that combines hard-label supervised cross-entropy loss with soft-label distillation loss. The cross-entropy loss acts as a regularizing "anchor," ensuring that the student model prioritizes predicting the ground-truth tokens. 

Formally, the cross-entropy loss is computed only over the output sequence tokens (where prompt tokens are masked with a label of $-100$ to exclude them from gradient computation):
\begin{equation}
    \mathcal{L}_{CE} = -\frac{1}{N_{\text{out}}} \sum_{t=1}^{N_{\text{out}}} \log P_{\text{student}}(y_t \mid y_{<t}, x),
\end{equation}
where $N_{\text{out}}$ is the number of output tokens, $y_t$ is the ground-truth token at step $t$, and $x$ represents the prompt context.

\subsection{Soft-Label Training Objective with Top-K Logits}

In addition to the hard-label anchor, the student model minimizes the divergence between its probability distribution and the soft-label distribution of the teacher model. The soft targets generated by the teacher contain rich, dark knowledge regarding the relationships between vocabulary tokens.

To optimize storage and computational efficiency, rather than computing the Kullback-Leibler (KL) divergence over the entire vocabulary space, we store only the top $K$ (where $K=50$) largest logits generated by the teacher model. Let $V_{\text{top}}$ represent the set of $K$ tokens with the highest logit values produced by the teacher at output step $t$. Both the teacher and student probability distributions are normalized over this restricted Top-K vocabulary space using a temperature parameter $T > 1$:
\begin{equation}
    P_{\text{teacher}}^T(v) = \frac{\exp(z_t^{\text{teacher}}(v) / T)}{\sum_{w \in V_{\text{top}}} \exp(z_t^{\text{teacher}}(w) / T)}, \quad \forall v \in V_{\text{top}},
\end{equation}
\begin{equation}
    P_{\text{student}}^T(v) = \frac{\exp(z_t^{\text{student}}(v) / T)}{\sum_{w \in V_{\text{top}}} \exp(z_t^{\text{student}}(w) / T)}, \quad \forall v \in V_{\text{top}},
\end{equation}
where $z_t(v)$ represents the logit value for token $v$ at step $t$.

The soft-label distillation loss $\mathcal{L}_{KD}$ is defined as the KL divergence between these temperature-scaled distributions, normalized by the sequence length $N_{\text{out}}$ and scaled by a factor of $T^2$ to preserve gradient magnitude:
\begin{equation}
    \mathcal{L}_{KD} = T^2 \cdot D_{KL}\left( P_{\text{teacher}}^T \parallel P_{\text{student}}^T \right) = T^2 \cdot \frac{1}{N_{\text{out}}} \sum_{t=1}^{N_{\text{out}}} \sum_{v \in V_{\text{top}}} P_{\text{teacher}}^T(v) \log \left( \frac{P_{\text{teacher}}^T(v)}{P_{\text{student}}^T(v)} \right).
\end{equation}
The final joint objective for Phase 1 is a weighted combination controlled by the hyperparameter $\alpha_1 \in [0, 1]$:
\begin{equation}
    \mathcal{L}_{\text{Phase 1}} = \alpha_1 \mathcal{L}_{KD} + (1 - \alpha_1) \mathcal{L}_{CE}.
\end{equation}
The complete computation flow of this phase, covering the parallel teacher--student forward passes and the combination of the two loss components, is depicted in Figure~\ref{fig:phase1_distillation}.

\input{figs/tikz/fig_3_2_offline_logit_kd}

\subsection{Regularization via Quantized Low-Rank Adaptation}

To perform offline logit distillation on resource-constrained hardware, we apply Quantized Low-Rank Adaptation (QLoRA) to the student model. QLoRA acts as an implicit regularization mechanism. During training, the base weights of the student model are frozen in a 4-bit NormalFloat (NF4) precision format, and trainable Low-Rank Adapter matrices (LoRA) are inserted parallel to the query, key, value, and projection layers.

By freezing the base pre-trained weights, QLoRA prevents the model from experiencing catastrophic forgetting --- a common failure mode where a model loses its general language comprehension capabilities during specialized domain adaptation. The low-rank updates are forced to capture only the essential delta required to align the student's output logits with the teacher's distribution, thereby stabilizing the gradient flow and reducing the risk of memorization on the static dataset.

\subsection{Hyperparameter Design and Justification}

The selection of training hyperparameters is critical to prevent catastrophic forgetting and stabilize the learning trajectory of the student model. Small Language Models (SLMs) are particularly vulnerable to gradient variance and model collapse when adapted using high learning rates. Consequently, rather than employing standard fine-tuning hyperparameters, we formulate a conservative training configuration designed to stabilize the adaptation process.

First, we restrict the learning rate to $2 \times 10^{-5}$ to prevent large weight updates from disrupting pre-trained parameters. Second, the distillation weight $\alpha_1$ is set to $0.3$, ensuring that the hard-label cross-entropy anchor $\mathcal{L}_{CE}$ exerts stronger guidance (with weight $1 - \alpha_1 = 0.7$) over the learning process. Third, we establish a larger effective batch size of 64 to reduce gradient variance and ensure a smooth, monotonic convergence of the joint loss. This design balances the alignment of specific legal vocabulary with the preservation of general linguistic capabilities.

