\section{Datasets}

To systematically train and evaluate the proposed legal reasoning models, we leverage a domain-specific dataset tailored for Vietnamese legal query answering. The datasets utilized throughout this research are built upon the foundation of the ViLawQA corpus, which is adapted from the Vietnamese \texttt{LegalSLM} shared task \parencite{le-etal-2025-overview}. 

Specifically, the raw data is compiled by acquiring three distinct datasets hosted on Hugging Face:
\begin{enumerate}
    \item \textbf{Multiple-Choice QA:} The \texttt{thangvip/vilawqa} repository containing 1,007 records for statutory option selection.
    \item \textbf{Natural Language Inference:} The \texttt{thangvip/vilawqa-nli} repository containing 1,421 records for legal document relevance evaluation.
    \item \textbf{Legal Syllogism:} The \texttt{thangvip/vilawqa-syllo} repository containing 1,000 records for complex deductive reasoning.
\end{enumerate}

These heterogeneous repositories are clone-retrieved and standardized into a unified schema using a custom converter module, resulting in 3,428 raw records. To prevent data leakage and preserve task distributions across folds, this consolidated corpus is partitioned into training (80\%), validation (10\%), and testing (10\%) splits. This partitioning is executed via stratified random sampling, utilizing the \texttt{task\_type} attribute as the stratification key to maintain identical task proportions across all divisions. After applying data cleaning and formatting filters, this process yields 2,603 high-quality samples for the training set and 325 samples for the internal test split.

Furthermore, to evaluate the models' generalization capabilities on an external benchmark, we employ an additional public evaluation set derived from the official public test set of the \texttt{VLSP2025-LegalSML} shared task \parencite{le-etal-2025-overview}.\footnote{The dataset is publicly accessible at \url{https://huggingface.co/datasets/VLSP2025-LegalSML/Public-Test}.} This public test set consists of 440 samples (146 Multiple-Choice, 150 Natural Language Inference, and 144 Syllogism) and has been converted into the ViLawQA format for evaluation. The dataset splits and evaluation sets utilized in this research are summarized in Table~\ref{tab:dataset_summary}.

\begin{table}[htbp]
\centering
\caption{Summary of Dataset Splits and Evaluation Sets}
\label{tab:dataset_summary}
\begin{tabular}{llcccc}
\hline
\textbf{Split} & \textbf{Source / Dataset} & \textbf{MCQ} & \textbf{NLI} & \textbf{Syllogism} & \textbf{Total} \\ \hline
Train & \texttt{train} & -- & -- & -- & 2,603 \\
Valid & \texttt{val} & -- & -- & -- & 343 \\
Test (Split) & \texttt{test} & 100 & 141 & 84 & 325 \\
Test (Public) & \href{https://huggingface.co/datasets/VLSP2025-LegalSML/Public-Test}{VLSP2025 Public-Test} & 146 & 150 & 144 & 440 \\ \hline
\end{tabular}
\end{table}

In the following subsections, we describe the composition of the supervised fine-tuning training dataset, the structure of the evaluation and test sets, and the construction of the preference alignment pairs for Direct Preference Optimization.

\subsection{Training Datasets}

The primary training dataset used for supervised fine-tuning and logit distillation is \texttt{vilawqa\_sft\_train.jsonl}, which comprises 2,603 high-quality, cleaned samples. Each instance in this dataset combines task directives and contextual legal inputs to guide the model's generation. Specifically, the input configuration is structured into two main fields:
\begin{itemize}
    \item \textbf{Instruction:} A task-specific directive defining the goal (e.g., instructing the model to answer a legal question based on the provided statutes).
    \item \textbf{Input:} A concatenated context containing raw statutory rules (e.g., specific clauses, decrees, or articles extracted from active Vietnamese legislation) and a factual query describing a concrete legal situation.
\end{itemize}

The corresponding target output (the ground-truth response) is designed to cultivate chain-of-thought (CoT) reasoning capabilities. Approximately 75\% of the training samples contain intermediate reasoning chains wrapped within HTML-style tags, specifically \texttt{<think> ... </think>}. These traces reflect the step-by-step thinking process generated by the pre-aligned teacher model, including premise extraction and rule matching. The remaining 25\% of the target outputs consist of direct answers or final legal verdicts, ensuring that the model is exposed to both reasoning-intensive and concise response patterns. A representative JSON format of a training sample is illustrated in the verbatim block below:

\begin{verbatim}
{
  "id": "vilawqa_syllo_00079",
  "task_type": "syllogism",
  "instruction": "Answer the legal question...",
  "input": "Legal basis:\nPoint c, Clause 1, Article 55...
            \nQuestion:\nReward amount for the medal...",
  "output": "<think>\nQuestion to answer:...\n
             Conclusion: Reward amount...\n</think>\n
             The reward amount..."
}
\end{verbatim}

\subsection{Evaluation and Test Sets}

For evaluation, we employ two test sets: (i) an internal test set, \texttt{vilawqa\_sft\_test.jsonl} (referred to as Split-Test), consisting of 325 balanced samples obtained from the 10\% stratified split of the unified ViLawQA corpus, and (ii) an external public test set (referred to as Public-Test) consisting of 440 samples from the \texttt{VLSP2025-LegalSML} shared task. To ensure a comprehensive assessment of the models' generalization capabilities across different legal reasoning modalities, both test sets are categorized into three core tasks:
\begin{enumerate}
    \item \textbf{Legal Natural Language Inference (NLI):} This task assesses the model's capability to determine the logical relevance of a given statutory context to a specific factual query. To establish clear evaluation boundaries, the inference labels are standardized into three unified categories: \texttt{relevant} (the statutory provision is directly relevant to answering the query), \texttt{not\_relevant} (the provision is irrelevant to the query), and \texttt{unknown} (the relationship cannot be determined based on the provided context).
    \item \textbf{Multiple-Choice Questions (MCQ):} This task evaluates statutory comprehension and decision-making by requiring the model to select the correct answer among four distinct options (\texttt{A}, \texttt{B}, \texttt{C}, or \texttt{D}) given a detailed legal situation and relevant legal bases.
    \item \textbf{Legal Syllogism:} This represents the most complex deductive task, testing the model's ability to perform multi-step reasoning. The model must apply a major premise (applicable statutes) to a minor premise (factual situation) to derive a logical conclusion. Evaluation on this task measures the ROUGE-L metric and Exact Match on the final answer to evaluate both the reasoning fluency and output precision.
\end{enumerate}

\subsection{Direct Preference Optimization Preference Data}

To align the student models and correct reasoning fallacies during the third training phase, we construct a preference dataset consisting of triplets $(x, y_w, y_l)$. Here, $x$ represents the prompt (the instruction and statutory input), $y_w$ denotes the chosen response (the ground-truth output), and $y_l$ represents the rejected response (an incorrect rollout generated by the student model). Preference pairs are synthesized by prompting the student models to generate rollouts on the training set and diagnosing their errors. 

We compare two extraction strategies for constructing these preference pairs:
\begin{itemize}
    \item \textbf{Rule-Based Extraction:} This approach utilizes regular expressions and ROUGE-L similarity thresholding to identify errors. While simple, this method is sensitive to syntactic variations and formatting inconsistencies. It extracted 2,313 preference pairs for the \texttt{Qwen3-0.6B-Base} model. However, it suffered from high noise, frequently misclassifying semantically sound responses as incorrect due to formatting deviation, or accepting incorrect answers that coincidentally matched the regular expression templates.
    \item \textbf{LLM Judge Extraction:} This method leverages a large language model (\texttt{GPT-4o-mini}) to perform a deep semantic evaluation of both the thinking process inside the \texttt{<think>} tags and the final verdict. This approach successfully filters out formatting anomalies while capturing logical errors. Consequently, it yields cleaner datasets: 2,065 high-quality pairs for the \texttt{Qwen3-0.6B-Base} model (\texttt{A2\_dpo.jsonl}) and 1,434 pairs for the \texttt{Qwen3-1.7B-Base} model (\texttt{A3\_dpo.jsonl}).
\end{itemize}
