\section{Results and Analysis}

This section reports and analyzes the empirical results of the proposed three-phase distillation and alignment framework. We organize the discussion around two complementary evaluation regimes: an in-distribution assessment on the 325-sample internal Split-Test, which isolates the contribution of each design choice, and an out-of-distribution assessment on the 440-sample VLSP2025 Public-Test, which probes generalization. Throughout this section, the \textit{Overall} score denotes the unweighted mean of the three task-level metrics (MCQ Accuracy, NLI Accuracy, and Syllogism ROUGE-L), and all reported figures are obtained under greedy decoding to ensure deterministic and reproducible evaluation. We note that this Overall score aggregates an accuracy metric with a lexical-overlap metric (ROUGE-L) of differing scales and interpretations; we therefore treat it only as a coarse summary indicator and ground our analysis primarily on the individual per-task metrics rather than on the composite value alone.

\subsection{Impact of Hyperparameter Tuning}

We first examine the effect of the optimization configuration by contrasting the initial distillation setup (V1) with the corrected setup (V2). In V1, an aggressive learning rate ($\eta = 1.0 \times 10^{-4}$) combined with a small effective batch size of 16 and a balancing weight of $\alpha = 0.5$ induced severe \textit{catastrophic forgetting}: the cross-entropy loss against the hard labels, rather than decreasing, rose monotonically during the later training steps (from approximately 0.3 to 0.7), indicating that the student was discarding its supervised fine-tuning (SFT) knowledge in order to chase the teacher distribution. This degradation manifested qualitatively as incoherent, code-switched generations mixing Vietnamese, English, and Chinese tokens.

The V2 configuration mitigates this instability by reducing the learning rate fivefold to $\eta = 2.0 \times 10^{-5}$, enlarging the effective batch size to 64 to suppress gradient noise, and increasing the relative weight of the hard-label objective ($\alpha = 0.3$, i.e., 70\% cross-entropy and 30\% distillation). Under V2, the cross-entropy loss converges smoothly from 0.5 to 0.25, preserving the model's original reasoning ability. As reported in Table~\ref{tab:hp_tuning}, this correction yields substantial improvements in the Overall score across all three configurations, most notably for the 1.7B model, which improves from 0.3644 (E3) to 0.5322 (A3).

\begin{table}[htbp]
\centering
\caption{Impact of Hyperparameter Tuning (V1 vs. V2) on the 325-sample Split-Test}
\label{tab:hp_tuning}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llccccccccc}
\hline
\textbf{Run} & \textbf{Method} & \textbf{Size} & \textbf{LR} & \textbf{$\alpha$} & \textbf{$T$} & \textbf{MC Acc} & \textbf{NLI Acc} & \textbf{NLI F1} & \textbf{Syl R-L} & \textbf{Overall} \\ \hline
E1 & LoRA V1  & 0.6B & 1.0e-4 & 0.5 & 2.0 & 0.2772 & 0.3475 & 0.3808 & 0.1623 & 0.2560 \\
A1 & LoRA V2  & 0.6B & 2.0e-5 & 0.3 & 1.5 & 0.4059 & 0.2270 & 0.2416 & 0.1741 & \textbf{0.4520} \\
E2 & QLoRA V1 & 0.6B & 1.0e-4 & 0.5 & 2.0 & 0.3564 & 0.4113 & 0.5322 & 0.2104 & 0.2536 \\
A2 & QLoRA V2 & 0.6B & 2.0e-5 & 0.3 & 1.5 & 0.4158 & 0.5886 & 0.4390 & 0.3804 & \textbf{0.4622} \\
E3 & LoRA V1  & 1.7B & 1.0e-4 & 0.5 & 2.0 & 0.5149 & 0.5957 & 0.7025 & 0.3627 & 0.3644 \\
A3 & LoRA V2  & 1.7B & 2.0e-5 & 0.3 & 1.5 & 0.4455 & 0.7305 & 0.5018 & 0.4208 & \textbf{0.5322} \\ \hline
\end{tabular}%
}
\end{table}

\subsection{Low-Rank Adaptation vs. Quantized Low-Rank Adaptation}

We next isolate the effect of the adaptation strategy by comparing standard Low-Rank Adaptation (LoRA) \parencite{hu2021loralowrankadaptationlarge} against Quantized Low-Rank Adaptation (QLoRA), under the identical V2 configuration on the 0.6B student (runs A1 and A2 in Table~\ref{tab:hp_tuning}). The pure-LoRA variant (A1) reaches its peak performance early, at approximately step 60, after which its score degrades continuously due to forgetting; correspondingly, its cross-entropy loss drifts upward (0.3 $\rightarrow$ 0.7).

In contrast, the QLoRA variant (A2) freezes the base weights in a 4-bit (NF4) representation \parencite{dettmers2022llmint88bitmatrixmultiplication} and confines updates to the adapter parameters. This quantization acts as a form of \textit{implicit regularization} that anchors the prior SFT knowledge, allowing the cross-entropy loss to decrease steadily (0.5 $\rightarrow$ 0.25) and the model to keep improving until step 120. The practical benefit is most pronounced on the most difficult task: QLoRA raises the Syllogism ROUGE-L from 0.1741 to 0.3804, a relative gain of roughly 33\% over LoRA, and improves the Overall score from 0.4520 to 0.4622. We therefore adopt QLoRA as the default adaptation strategy for the 0.6B student in all subsequent phases.

\subsection{Model Size Comparison: 0.6 Billion vs. 1.7 Billion Parameters}

Parameter capacity proves decisive for deep, multi-step legal reasoning. To quantify this gap independently of the final test metrics, we apply an LLM-as-a-Judge protocol (\texttt{GPT-4o-mini}) to the responses generated by the two distilled students on the training set, categorizing each response into \texttt{OK}, \texttt{RISKY}, \texttt{PARTIAL}, or \texttt{WRONG}. The 1.7B model (A3) attains a fully acceptable (\texttt{OK}) rate of 44.9\%, more than double the 20.7\% achieved by the 0.6B model (A2). The disparity is sharpest on the Syllogism task, where the \texttt{OK} rate rises from only 6\% for the 0.6B student to 20\% for the 1.7B student, reflecting the larger model's superior capacity to process long statutory contexts and sustain coherent deductive chains. This trend, in which the 1.7B model consistently dominates the 0.6B model, recurs across every subsequent phase of the pipeline.

\subsection{Supervised Fine-Tuning vs. Offline Knowledge Distillation}

We now assess the contribution of offline logit distillation \parencite{hinton2015distillingknowledgeneuralnetwork} relative to the SFT baselines, using the progression reported in Table~\ref{tab:progression}. Rather than claiming a uniform improvement, we analyze the effect on a per-task basis, since the baselines and the distilled models exhibit complementary strengths. The 0.6B SFT baseline (B2) is already competent at NLI and free-form reasoning (NLI Accuracy 0.7160, Syllogism ROUGE-L 0.4280) but is weak at multiple-choice selection (MC Accuracy 0.1400). Offline distillation transfers the teacher's soft probability distribution and corrects precisely this deficiency, raising the MC Accuracy to 0.4158 and lifting the Overall score from 0.4282 to 0.4622. Symmetrically, the 1.7B SFT baseline (B3) is strong at MCQ (0.5000) but weak at syllogistic reasoning (ROUGE-L 0.2070); here distillation more than doubles the Syllogism ROUGE-L to 0.4208 and improves the Overall score from 0.4840 to 0.5322. These results confirm that the soft labels supply task-specific corrective signal beyond what the hard labels alone provide.

\begin{table}[htbp]
\centering
\caption{Performance Progression Across Training Phases on the 325-sample Split-Test}
\label{tab:progression}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llcccccc}
\hline
\textbf{Run} & \textbf{Stage} & \textbf{Size} & \textbf{MC Acc} & \textbf{NLI Acc} & \textbf{NLI F1} & \textbf{Syl R-L} & \textbf{Overall} \\ \hline
B2 & SFT (baseline)   & 0.6B & 0.1400 & 0.7160 & 0.4730 & 0.4280 & 0.4282 \\
A2 & Offline KD (V2)  & 0.6B & 0.4158 & 0.5886 & 0.4390 & 0.3804 & 0.4622 \\
A2 & On-Policy KD     & 0.6B & 0.1600 & 0.4960 & 0.3800 & 0.2750 & 0.3104 \\
A2 & DPO Alignment    & 0.6B & 0.1000 & 0.3191 & 0.2387 & 0.1392 & 0.1861 \\ \hline
B3 & SFT (baseline)   & 1.7B & 0.5000 & 0.7450 & 0.5040 & 0.2070 & 0.4840 \\
A3 & Offline KD (V2)  & 1.7B & 0.4455 & 0.7305 & 0.5018 & 0.4208 & 0.5322 \\
A3 & On-Policy KD     & 1.7B & 0.3600 & 0.7870 & 0.5310 & 0.3750 & 0.5075 \\
A3 & DPO Alignment    & 1.7B & 0.2000 & 0.8156 & 0.8333 & 0.5684 & \textbf{0.5280} \\ \hline
\end{tabular}%
}
\end{table}

\subsection{Offline vs. On-Policy Knowledge Distillation}

The second phase addresses \textit{exposure bias}, the train-test discrepancy that arises when a student is trained exclusively on ground-truth prefixes yet must condition on its own generations at inference time \parencite{agarwal2024onpolicydistillationlanguagemodels, song2026surveyonpolicydistillationlarge}. On-policy distillation closes this gap by letting the student generate its own rollouts and learn from teacher feedback on those self-generated trajectories. For the 1.7B student (A3), this mechanism delivers a marked gain on the reasoning-classification task, pushing the NLI Accuracy from 0.7305 to 0.7870 (Table~\ref{tab:progression}). The slight decrease in the Overall score (0.5322 $\rightarrow$ 0.5075) is attributable to the token-level on-policy objective penalizing the free-form reasoning tokens, which marginally lowers the Syllogism ROUGE-L.

For the 0.6B student (A2), however, on-policy distillation is counterproductive: its limited parameter capacity causes the self-generated rollouts to accumulate excessive noise, leading to a logical collapse in which the Overall score falls to 0.3104. This contrast reinforces the capacity argument of the preceding subsection: the benefit of learning from self-generated trajectories is contingent on the student being large enough to produce trajectories of usable quality.

\subsection{Direct Preference Optimization Alignment Impact}

In the final phase, the distilled students are aligned with Direct Preference Optimization (DPO) \parencite{rafailov2024directpreferenceoptimizationlanguage} using diagnosis-driven preference pairs. For the 1.7B model, DPO substantially reshapes the task profile rather than uniformly improving it: NLI Accuracy rises from 0.7305 to 0.8156 and Syllogism ROUGE-L surges from 0.4208 to 0.5684, while MCQ Accuracy drops from 0.4455 to 0.2000 (Table~\ref{tab:progression}). Because these gains and losses partly offset, the Overall score of A3\_dpo (0.5280) is marginally below that of the offline-distilled checkpoint A3\_offline\_kd (0.5322); DPO thus does not improve the aggregate metric, but it yields the strongest NLI and syllogistic-reasoning performance of any in-distribution configuration. By contrasting correct reasoning traces against the model's own diagnosed errors, DPO refines residual logical blind spots and enforces stricter adherence to the structured \texttt{<think>} reasoning format, at the expense of multiple-choice accuracy.

In sharp contrast, applying DPO to the 0.6B model is detrimental, collapsing its Overall score to 0.1861. We attribute this to the probability-displacement penalty of the DPO objective being too severe for a sub-billion-parameter network to absorb without destabilizing its already-fragile reasoning behavior. Consequently, DPO is beneficial only at the 1.7B scale, and the 0.6B alignment result stands as a cautionary finding regarding the interaction between preference optimization and model capacity.

\subsection{Generalization on the Public Test Set}

The preceding analysis characterizes in-distribution behavior. To assess generalization, we evaluate all eight models on the 440-sample VLSP2025 Public-Test, an out-of-distribution benchmark drawn from the official shared task \parencite{le-etal-2025-overview}. The results, reported in Table~\ref{tab:vlsp_ood}, reveal a notable reversal relative to the in-distribution ranking and warrant a critical discussion of the proposed pipeline.\footnote{The official VLSP2025 protocol employs an LLM-as-a-Judge for the Syllogism task; we instead report ROUGE-L as an internal proxy metric to enable consistent comparison across all models, so these figures are not the official competition scores.}

\begin{table}[htbp]
\centering
\caption{Generalization Results on the 440-sample VLSP2025 Public-Test (Out-of-Distribution)}
\label{tab:vlsp_ood}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llccccccc}
\hline
\textbf{Run} & \textbf{Stage} & \textbf{Size} & \textbf{MC Acc} & \textbf{NLI Acc} & \textbf{NLI F1} & \textbf{Syl R-1} & \textbf{Syl R-L} & \textbf{Overall} \\ \hline
B2 & SFT          & 0.6B & 0.541 & 0.707 & 0.480 & 0.449 & 0.266 & 0.5047 \\
B3 & SFT          & 1.7B & \textbf{0.740} & 0.873 & 0.590 & 0.581 & 0.305 & \textbf{0.6395} \\
A2 & KD V2        & 0.6B & 0.199 & 0.607 & 0.428 & 0.610 & 0.331 & 0.3787 \\
A3 & KD V2        & 1.7B & 0.219 & 0.720 & 0.525 & 0.667 & 0.353 & 0.4307 \\
A2 & KD + DPO     & 0.6B & 0.151 & 0.207 & 0.177 & 0.165 & 0.122 & 0.1599 \\
A3 & KD + DPO     & 1.7B & 0.212 & 0.893 & \textbf{0.897} & 0.409 & 0.282 & 0.4627 \\
A2 & On-Policy KD & 0.6B & 0.055 & 0.673 & 0.482 & 0.609 & 0.313 & 0.3471 \\
A3 & On-Policy KD & 1.7B & 0.555 & \textbf{0.920} & 0.617 & \textbf{0.679} & \textbf{0.372} & 0.6155 \\ \hline
\end{tabular}%
}
\end{table}

Three findings emerge. First, the 1.7B SFT baseline (B3) achieves the highest Overall score on the Public-Test (0.6395), surpassing all distilled and aligned variants. This indicates that knowledge distillation, while effective on in-distribution data, over-specializes the student toward the training distribution at the expense of generalization, a tendency most visible on the MCQ task. Second, among the distilled models, the 1.7B On-Policy KD checkpoint (A3) is the strongest generalizer (Overall 0.6155, second overall), attaining both the highest NLI Accuracy (0.920) and the highest Syllogism ROUGE-L (0.372) in the entire table; this demonstrates that breaking exposure bias confers a generalization advantage that persists on unseen data. Third, MCQ accuracy is a systematic weakness of every distilled model (all below 0.56, versus 0.74 for the 1.7B SFT baseline), and the 0.6B On-Policy model nearly loses the capability altogether (0.055).

Taken together, the in-distribution and out-of-distribution results yield two consistent conclusions. Model scale is the dominant factor: the 1.7B family outperforms its 0.6B counterpart across every pipeline stage, frequently by a factor of $1.5$–$2\times$. Moreover, the optimal checkpoint is contingent on both the evaluation regime and the target task: in-distribution, A3\_offline\_kd attains the highest Overall score while A3\_dpo dominates the NLI and syllogistic-reasoning tasks; under distribution shift, the 1.7B On-Policy KD model is the most robust distilled configuration. We therefore refrain from designating a single universally optimal model and instead report the trade-off explicitly.
