\section{Qualitative Analysis}

Beyond the aggregate metrics reported in the preceding sections, we conduct a qualitative analysis to characterize the concrete behavioral changes induced by the distillation pipeline. We first compare model outputs on a small, balanced set of representative cases, and then catalogue the recurring error patterns observed before and after distillation.

\subsection{Nine-Sample Inference Comparison}

We select nine representative test cases, comprising three Natural Language Inference (NLI), three Multiple-Choice (MCQ), and three Legal Syllogism instances, and inspect the generations of four models under greedy decoding: the SFT baseline and the distilled (KD) checkpoint for each of the 0.6B and 1.7B students. The number of correct responses per task is summarized in Table~\ref{tab:nine_sample}.

\begin{table}[htbp]
\centering
\caption{Nine-Sample Inference Comparison (Correct Responses per Task, Greedy Decoding)}
\label{tab:nine_sample}
\begin{tabular}{lcccc}
\hline
\textbf{Model} & \textbf{NLI (3)} & \textbf{MCQ (3)} & \textbf{Syllogism (3)} & \textbf{Total} \\ \hline
0.6B SFT & 2/3 & 2/3 & 0/3 & 4/9 \\
0.6B KD  & 3/3 & 2/3 & 0/3 & 5/9 \\
1.7B SFT & 3/3 & 1/3 & 1/3 & 5/9 \\
1.7B KD  & 3/3 & 2/3 & 1/3 & \textbf{6/9} \\ \hline
\end{tabular}
\end{table}

Two complementary effects are evident. First, knowledge distillation acts as a targeted error-correction mechanism: for the 0.6B student, distillation corrects an erroneous NLI verdict produced by the SFT baseline (raising NLI from 2/3 to 3/3), while for the 1.7B student it corrects an incorrect MCQ selection (raising MCQ from 1/3 to 2/3). In both cases the soft labels supplied by the teacher patch a specific reasoning gap inherited from supervised fine-tuning. Second, model scale governs the ceiling on deductive capability: the 0.6B student fails all three Syllogism cases (0/3) regardless of distillation, whereas the 1.7B student solves one of them and correctly extracts the relevant statutory figures. Distillation thus functions as a corrective catalyst that transfers knowledge from the 4B teacher, but it cannot, on its own, compensate for the capacity limitations of the smallest student.

\subsection{Error Patterns Before and After Distillation}

A qualitative inspection of the failure modes further clarifies how the pipeline reshapes model behavior. Prior to distillation, the SFT student models exhibit two characteristic pathologies. The first is degenerate repetition, in which the model enters an infinite loop emitting the same token sequence. The second is unintended code-switching: when confronted with complex legal terminology, the student abruptly drifts from Vietnamese into English or Chinese in the middle of a reasoning chain, breaking both coherence and the expected output language.

After distillation and DPO alignment, these pathologies are substantially suppressed. The student maintains coherent, monolingual Vietnamese output throughout the reasoning chain, and it adheres reliably to the structured \texttt{<think> ... </think>} format, articulating its reasoning before committing to a final verdict. The aligned model also improves the precision with which it cites the applicable statutory identifiers (specific articles and clauses) within a concrete legal scenario. These observations are consistent with the quantitative gains reported in Section~4.3: the behavioral shift from incoherent, code-switched generation toward disciplined, format-compliant Vietnamese reasoning is the qualitative counterpart of the measured improvements on the NLI and Syllogism tasks.
